---
title: "Using Covidestim"
author: "Melanie H. Chitwood, Marcus Russi" 
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Covidestim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette demonstrates running the model on NYC data and viewing the result.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(covidestim)
```

## Running the model 
### Data

We will demonstrate how to use the covidestim package using data from the early
NYC epidemic that the package makes available through the `example_nyc_data()`
function.

Users can use case and/or death data with the model. Each data type should be
entered as its own dataframe with two columns: `date` and `observation`.

```{r}
head(example_nyc_data('cases'))
head(example_nyc_data('deaths'))
```

### Generate a model configuration 

Input data and model run options are summarized into a covidestim configuration with the function call `?covidestim`. Users must specify the number of days of data in the input data. More information on data input can be accessed at `?input_cases` and `?input_deaths`. Printing the result of `covidestim()` will produce a summary of model run conditions, priors, and data inputs. 

Input data and model configuration options are rolled up into a model configuration object as follows:

```{r}
cases  <- example_nyc_data('cases')
deaths <- example_nyc_data('deaths')

cfg <- covidestim(ndays = nrow(cases), region = 'New York') +
  input_cases(cases) +
  input_deaths(deaths)

print(cfg)
```

There are a few possible variations on this:

1. Custom priors can be set by adding them, just as `input_cases()` and
   `input_deaths()` were added.

2. A population size can be set, to make the model aware of an upper bound on
   how many individuals could be infected.

3. `input_cases()` and `input_deaths()` can accept a second parameter, `type`, to
   specify if the data is organized by date-of-report, or date-of-event.

If you set custom priors, be sure to check the output of `print(cfg)` to see
that your prior is present in the configuration.

### Initializing a model run

Finally, the model is run with the function `?run.covidestim`. Don't forget to
assign the a name to the output of the run - you could lose it!. Runtimes vary
depending on the size and quality of the data, and may take over an hour. If
you are running the model in an interactive environment like RStudio, periodic
status updates will appear in the console. Here, we will use the BFGS algorithm
to quickly produce a result, as NUTS, the method used by `run()`, takes longer.

```{r}
result <- runOptimizer(cfg)
```

Typically, BFGS will occasionally fail to converge. This is what causes the
non-zero return codes. `runOptimizer.covidestim()` will automatically exclude these, and 
select the BFGS run that maximizes the posterior.

## Accessing model output 

Calling `run()` or `runOptimizer()` on a valid covidestim configuration will
produce an S3 object of class `covidestim_result`. Here, we describe the summary
functions in covidestim to view model results. In the vignette `vignette("results")`,
we describe the various objects within `covidestim_result`. However, most users
will likely find that the `summary.covidestim_result()` function provides all the information
they need.

### Summarizing covidestim output

A covidestim object can be easily converted to a summary data frame with the
`?summary.covidestim_result` function. Each row of the data frame is a date. Each
column of the data frame is a value of interest. Variables appened with .hi or
.lo indicate the 0.975 and 0.025 quantiles, respectively.

```{r}
result.summary <- summary(result)
print(result.summary[40:45, 1:5])
```

## Visualizing model output

Model results can be visualized with `?viz`. Two figures will be produced.

```{r}
plots <- viz(result)
```

The figure 'observedVsFitted' plots the observed cases and deaths (the input
data) against the modeled values for reported cases and deaths. Because `runOptimzier.covidestim_result()`
does not have CI's, there will be a warning relating to the `NA`-valued CI columns
that are present in the output.

```{r}
plots$observedVsFitted
```

The figure 'infectionsAndCases' plots modeled reported cases against modeled
'true' new infections. The latter is what we estimate to be the complete,
un-lagged time series of incident infections that would give rise to the
trends we observe in the case and death data. 

```{r}
plots$infectionsAndCases
```

